Exploratory data structure comparisons using Principal Component Analysis

### 2017-03-15 article/article_v0.01.tex
Slides + added comments and minor corrections to the rest of section 3.

%\title{Something that is not PCADSC}
%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
\title{Explorative data structure comparison with application to \ldots}

@@

Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}. Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers. The associated \emph{rank-q-reconstruction error} is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
\emph{Principal component analysis} (PCA) ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero \hl{Anne: What about covariance matrices that are not of full rank (e.g. because of colinearity)? Won't we find eigenvalue multiplicities $> 1$ here as well?}. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
\hl{Anne: Should we maybe assume standardized $x$s here also? I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason.} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.

@@

Before describing the details of the  \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot} we make some general remarks. First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data. Therefore, the variables should always be scaled prior to performing PCA. Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices. \hl{Do we still need more about this topic? Then we can add something like:
\begin{itemize}
\item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \V(u^T X^T) $$
\item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$. Note that $(1, 0, ..., 0) X^T = x_i$. Then, clearly a vector $u$ that only picks out the first variable from the data matrix $X$, namely $u^T = (1, 0, ..., 0)$ will allow us to only look at the variance of $x_1$, which is very large. This has nothing to do with the interplay between variables, as the phenomenon will disappear if we scale $x_1$
\item A more formal explanation could come from this kind of intuition, if needed.
\end{itemize}
}
The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1. This also holds for the reallocated datasets, and in particular we have $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$. The random reallocations in step 4 provide resamples under the null hypothesis that the two sample populations have the same structure in the $d$ variables, except that the two populations are allowed to have the own standard deviations for each of the variables. This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2. These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a $p$-value for the null hypothesis may be computed. Such $p$-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).

@@

In step 3 we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
\end{align*}
\hl{Maybe more here on what the null hypothesis of these tests is?} 
From the random reallocations in step 4 we get $p$-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests. Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95 \% coverage intervals. If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.

@@

\subsubsection{Description of the \emph{HairPlot}} 
This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j} \eta_{1k}^\top \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
\end{align*}

The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using anticlockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset. If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length. Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells. The name \emph{HairPlot} arise by seeing the arrows as \emph{hairs}. Using this interpretation we may say that a more ``hairy'' plot indicates less agreement between the two loading patterns.

Along with the visual comparison we have also implemented numerical tests in step 3. Separately for the off-diagonal grid cells we may use the lengths $\mu_{jk}$ and $\nu_{jk}$ as test statistics, and for the on-diagonal grid cells we use the distance between the scaled loadings as test statistics, that is
\begin{equation*}
\chi_k = \lambda_{\max}^{-1} \lvert \lambda_{1k} \eta_{1k} - \lambda_{2k} \eta_{2k} \rvert, \quad \text{for $k=1,\dotsc,d$.}
\end{equation*}
In total this gives $2d^2 - d$ tests, which should be corrected for the family wise type I error using a methods that doesn't assume independence among the tests. We also construct overall Kolmogorov-Smirnov and Cram\'er-von Mises type test statistics by
\begin{align*}
\text{KS}_1 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k), &
\text{CvM}_1 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k)^2, \\
\text{KS}_2 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k), &
\text{CvM}_2 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k)^2.
\end{align*}

@@

\subsubsection{Description of the \emph{PancakePlot}}
This plot compares the loadings. The \emph{PancakePlot} consists of two parallel vertical barplots, one for each of the datasets, with one bar for each of the $d$ loading vector $\eta_{ij} \in \RR^d$. In the bars the relative contributions of the $d$ variables are visualized by separate colors for the different variables. Due to the indeterminacy of the sign all the signs are removed from the coefficients in the loadings, and the relative contribution is defined as the resulting coefficients divided by their total sum. The bars are ordered according to the variation components, and we also insert the cumulative variation components next to the barplots. The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.  \hl{Refer to example/show plot.}

@@

>>>>

%\title{Something that is not PCADSC}
%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
\title{Exploratory data structure comparisons using Principal Component Analysis}

@@

Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}. Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers. The associated \emph{rank-q-reconstruction error} is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
\emph{Principal component analysis} (PCA) ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero \hl{Anne: What about covariance matrices that are not of full rank (e.g. because of colinearity)? Won't we find eigenvalue multiplicities $> 1$ here as well?}. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
\hl{Anne: Should we maybe assume standardized $x$s here also? I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason. Also: Something is wrong with the dimensions here. $u$ is $d \times 1$, $x_i$ is $d \times 1$(?) - how do we get $\eta_j$ to also be $d \times 1$?} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.

@@

Before describing the details of the  \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot} we make some general remarks. First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data. Therefore, the variables should always be scaled prior to performing PCA. Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices. \hl{Do we still need more about this topic? Then we can add something like:
\begin{itemize}
\item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \tilde{\V}(u^T X^T) $$
\item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$. Note that $(1, 0, ..., 0) X^T = x_i$. Then, clearly a vector $u$ that only picks out the first variable from the data matrix $X$, namely $u^T = (1, 0, ..., 0)$ will allow us to only look at the variance of $x_1$, which is very large. This has nothing to do with the interplay between variables, as the phenomenon will disappear if we scale $x_1$
\item A more formal explanation could come from this kind of intuition, if needed.
\end{itemize}
}
The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1. This also holds for the reallocated datasets, and in particular we have $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$. The random reallocations in step 4 provide resamples under the null hypothesis that the two sample populations have the same structure in the $d$ variables, except that the two populations are allowed to have the own standard deviations for each of the variables. This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2. These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a $p$-value for the null hypothesis may be computed. Such $p$-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).

@@

In step 3 we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
\end{align*}
\hl{Maybe more here on what the null hypothesis and range of these test statistics are?} 
From the random reallocations in step 4 we get $p$-values for the Kolmogorov-Smirnov and the Cram\'er-von Mises Goodness-of-Fit tests. Step 4 is also used to visualize the sampling variation in a display, where we plot the observed curve together with 20 of the resampled curves as well as a shaded region visualizing pointwise 95 \% coverage intervals. If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.

@@

\subsubsection{Description of the \emph{HairPlot}} 
This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
%\begin{equation*}
%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
%= \lambda_{\max} \sum_{k=1}^d
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
%\end{equation*}
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset \hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?}. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
\end{align*}
\hl{Maybe emphasize inner product/angle-interpretation here?}
The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using anticlockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component for the first dataset in the coordinate system of the second dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset. \hl{Can we expand on this in a non-technical way?} If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length. Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells. The name \emph{HairPlot} arise by seeing the arrows as \emph{hairs}. Using this interpretation we may say that a more ``hairy'' plot indicates less agreement between the two loading patterns.

Along with the visual comparison we have also implemented numerical tests in step 3 \hl{Anne: Is a test really necessary here? Isn't the point of this plot to point us towards the most interesting components, not to quantify exactly how much they differ? I'm unsure as to whether a significant difference will necessarily be empirically/clinically relevant and likewise, if a non-significant difference will always be one we can disregard of...}. Separately for the off-diagonal grid cells we may use the lengths $\mu_{jk}$ and $\nu_{jk}$ as test statistics, and for the on-diagonal grid cells we use the distance between the scaled loadings as test statistics, that is
\begin{equation*}
\chi_k = \lambda_{\max}^{-1} \lvert \lambda_{1k} \eta_{1k} - \lambda_{2k} \eta_{2k} \rvert, \quad \text{for $k=1,\dotsc,d$.}
\end{equation*}
In total this gives $2d^2 - d$ tests, which should be corrected for the family wise type I error using a methods that doesn't assume independence among the tests. We also construct overall Kolmogorov-Smirnov and Cram\'er-von Mises type test statistics by
\begin{align*}
\text{KS}_1 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k), &
\text{CvM}_1 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \mu_{jk} + 1_{j=k} \chi_k)^2, \\
\text{KS}_2 &= \max_{j,k=1,\dotsc,d} (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k), &
\text{CvM}_2 &= \sum_{j=1}^d \sum_{k=1}^d (1_{j \neq k} \nu_{jk} + 1_{j=k} \chi_k)^2.
\end{align*}

@@

\subsubsection{Description of the \emph{PancakePlot}}
This plot compares the loadings. The \emph{PancakePlot} consists of two parallel vertical barplots, one for each of the datasets, with one bar for each of the $d$ loading vector $\eta_{ij} \in \RR^d$. In the bars the relative contributions of the $d$ variables are visualized by separate colors for the different variables. Due to the indeterminacy of the sign all the signs are removed from the coefficients in the loadings, and the relative contribution is defined as the resulting coefficients divided by their total sum. The bars are ordered according to the variation components, and we also insert the cumulative variation components next to the barplots. The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions.  \hl{Refer to example/show plot. Also, from Anne to Anne: A bit further elaboration here, especially mention the loadings as weights interpretation}

@@

\hl{Mention somewere how we propose for people to use the plots: First CE, then hair, then pancake focusing in hairy components.}
